Estimating the genome-wide contribution of selection to temporal allele frequency change

### 2019-09-24 manuscript/manuscript.tex
ms updates

\title{Estimating the genome-wide contribution of selection to temporal allele frequency change.}

@@

Additionally, we looked for a signal of temporal autocovariance in
\textcite{Bergland2014-ij}, a study that collected \emph{Drosophila
melanogaster} through Spring-Fall seasons across three years. With a pattern of
genome-wide fluctuating selection, we might expect a pattern of positive
covariances between similar seasonal changes, and negative covariances between
dissimilar seasonal changes. However, we find no such signal; we discuss this
in more depth in Supplementary Materials Section \ref{bergland-reanalysis}.

@@

%since
%drift also acts to inflate the magnitude of allele frequency change. By
%flipping the sign at the block-level, we

%we can  % (mean \gc{SNP} perwindow 4,606, median 4,830)
%analyze the genome-wide distribution of tile covariances for changes through
%time. We would like to compare the distribution of these tile covariances to a
%neutral expectation, but using a theoretic neutral null distribution would be
%inappropriate given complex dependencies across the genome due to linkage
%disequilibria (which can occur over long physical distances in E\&R and
%selection studies, \cite{Nuzhdin2013-gf,Baldwin-Brown2014-cl}).

%Averaging over such large numbers of loci is necessary to detect
%a signal of covariance in allele frequency changes when both drift (inflated in
%evolve-and-sequence studies which use small population sizes) and sampling
%variance can create spurious covariances.

%This could be caused by advantageous fitness
%backgrounds later becoming disadvantageous, either due to frequency-dependent
%selection, recombination breaking up coupled beneficial and deleterious
%alleles, or an environmental change in the direction of selection.
%However, the
%signal of a change in the direction of selection localized to a particular
%region of the genome may be washed out when we calculate genome-wide temporal
%covariances.

%Using the \textcite{Barghi2019-qy} study's data, we used the empirical neutral
%null to investigate whether there was a general trend for the temporal
%covariance $\cov(\Delta p_t, \Delta p_{t+k})$ to be positive for small $k$ (as
%expected under selection), but then become systematically negative for larger
%$k$ (as would occur under evolutionary processes that reverse the fitness of
%regions across the genome). Such a change in the fitness direction of
%haplotypes would not be discernible from genome-wide covariances like those in
%Figure \ref{fig:figure-1}.

% \begin{figure}[!ht]
%   \centering
%   \includegraphics[width=\textwidth]{}
%   \caption{}
%   \label{}
% \end{figure}

% \section{Materials and Methods}

% \subsection{Bias Correction}
% \subsection{}

@@

We would like to thank the authors of the original studies we've analyzed,
including Christian Schl{\"o}tterer, John Kelly, Kimberly Hughes, Frank Chan,
Campbell Rolian, Nick Barton, Alan Bergland, and Dmitri Petrov.  We would also
like to thank Doc Edge for helpful statistical advice, and Matt Osmond for
helpful discussions.

@@

\section{Supplementary Figures}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[]{figures/barghi-tailprobs-panels.pdf}
  \caption{}
  \label{suppfig:barghi-tailprobs-panels}
\end{figure}

@@

We also applied our temporal covariance approach to \textcite{Bergland2014-ij},
which found evidence of genome-wide fluctuating selection between Spring and
Fall seasons across three years \emph{Drosophila melanogaster}. As described in
\textcite{Buffalo2019-io}, we might expect positive covariances between like
seasons changes (i.e. Spring 2010 to Fall 2010 and Spring 2011 to Fall 2011),
and negative covariances between dislike seasonal changes (i.e. Fall 2009 to
Spring 2010 and Fall 2010 to Spring 2011) when fluctuating selection is strong
and acts genome-wide, as the original study found. However, while we find
temporal covariances that are non-zero, we find only weak support for a
seasonal fluctuating model driving these covariances.  In Supplementary Figure
\ref{suppfig:bergland-covs-figure}, we show the temporal covariances from
varying reference generations, across seasonal transitions that are alike (e.g.
the covariance between the allele frequency changes between Fall 2009 and
Spring 2009, and frequency changes between Fall 2010 and Spring 2010), and
dislike (e.g. the covariance between the allele frequency change between Fall
2009 and Spring 2009, and the frequency changes between Spring 2010 and Fall
2009). The first row of temporal covariance matrix is consistent with
fluctuating selection operating for two timepoints, as the first covariance is
negative, and the second is positive, and later covariances are not
statistically differentiable from zero (which could occur if LD and additive
genetic variance decay). However, the all other temporal covariances do not fit
the pattern we would expect under genome-wide fluctuating selection.

@@

  \caption{A: Scatterplot of the original unadjusted p-values from
    \textcite{Bergland2014-ij} and the p-values from our reanalysis of the same
    data using the same statistical methods; the minor discrepancy is likely due to software
    version differences. B: The histograms of the p-values of our reanalysis
    and the original \textcite{Bergland2014-ij} data; again the minor
    discrepancy is likely due to software differences. Overall, our
    implementation of Bergland et al.'s statistical methods produces results 
    very close to the original analysis.}

@@

It is unclear how strong the fluctuations would have to be to generate a
genome-wide average signal of fluctuating selection from temporal covariances.
For example, many loci could still show a signal of fluctuating selection, but
the average signal could be overwhelmed by other signals of other selection. To
investigate whether there was an genome-wide excess of loci showing evidence of
fluctuating selection we reanalyzed the data of \parencite{Bergland2014-ij}
using the same seasonal fluctuating model as the original paper. This model is
a Binomial logit-linked GLM fit per-locus, where the Spring/Fall seasons are
encoded as a dummy variable, and are regressed on the frequency data. We use
the same binomial weighting procedure as \textcite{Bergland2014-ij}, where the
weights are determined by the effective number of chromosomes, $N_{eff} = (2
n_t d_t - 1) / (2 n_t + d_t)$ ($n_t$ and $d_t$ are the number of diploid
individuals and the read depth at timepoint $t$, respectively). We fit this
model on all loci marked as used in the VCF provided with the
\textcite{Bergland2014-ij} study (doi:10.5061/dryad.v883p). Overall, our
p-values for the Wald test for each locus closely match those of the original
paper (Pearson correlation coefficient 0.98, p-value < $2.2 \times 10^{-16}$;
see Supplementary Figure \ref{suppfig:bergland-pvalue-comparison} A), and the
histograms of the p-values are nearly identical (Supplementary Figure
\ref{suppfig:bergland-pvalue-comparison} B). \textcite{Bergland2014-ij} find
loci with a significant association with season after a Benjamini and Hochberg
FDR p-value adjustment \parencite{Benjamini1995-jy}, however, the null
hypothesis of the Wald test does not give us an idea of the expected number of
variants that may spuriously fit the pattern of seasonal fluctuating
selection.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth]{figures/bergland-combined-hists.pdf}
  \caption{}
  \label{suppfig:bergland-pvalue-hist}
\end{figure}

>>>>

\title{Estimating the genome-wide contribution of selection to temporal allele frequency change}

@@

Additionally, we looked for a signal of temporal autocovariance in
\textcite{Bergland2014-ij}, a study that collected \emph{Drosophila
melanogaster} through Spring-Fall season pairs across three years. If there was
a strong  pattern of genome-wide fluctuating selection, we might expect a
pattern of positive covariances between similar seasonal changes, e.g.
Spring-Fall in two adjacent years, and negative covariances between dissimilar
seasonal changes, e.g. Spring-Fall and Fall-Spring in two adjacent years.
However, we find no such signal over years; we discuss this in more depth in
Supplementary Materials Section \ref{bergland-reanalysis}.

@@

@@

We would like to thank the authors of the original studies we've analyzed,
including Neda Barghi, Christian Schl{\"o}tterer, John Kelly, Kimberly Hughes,
Frank Chan, Campbell Rolian, Nick Barton, Alan Bergland, and Dmitri Petrov.  We
would also like to thank Doc Edge for helpful statistical advice, and Matt
Osmond for helpful discussions.

@@

\clearpage
\section{Supplementary Figures}

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[]{figures/barghi-tailprobs-panels.pdf}

  \caption{\textcite{Barghi2019-qy} tail probabilities compared to
  sign-permuted empirical null distribution for various $\alpha$ levels.}

  \label{suppfig:barghi-tailprobs-panels}
\end{figure}

@@

We also applied our temporal covariance approach to \textcite{Bergland2014-ij},
which found evidence of genome-wide fluctuating selection between Spring and
Fall seasons across three years \emph{Drosophila melanogaster}. As described in
\textcite{Buffalo2019-io}, if fluctuating selection pressure among time-periods
are the dominant genome-wide pattern, we might expect positive covariances
between like seasons changes (e.g. Spring 2010 to Fall 2010 and Spring
2011 to Fall 2011), and negative covariances between dislike seasonal changes
(e.g. Fall 2009 to Spring 2010 and Fall 2010 to Spring 2011). However,
while we find temporal covariances that are non-zero, we find only weak support
for a seasonal fluctuating model driving these covariances. In Supplementary
Figure \ref{suppfig:bergland-covs-figure}, we show the temporal covariances
from varying reference generations, across seasonal transitions that are alike
(e.g.  the covariance between the allele frequency changes between Fall 2009
and Spring 2009, and frequency changes between Fall 2010 and Spring 2010), and
dislike (e.g. the covariance between the allele frequency change between Fall
2009 and Spring 2009, and the frequency changes between Spring 2010 and Fall
2009). The first row of temporal covariance matrix is consistent with
fluctuating selection operating for two timepoints, as the first covariance is
negative, and the second is positive, and later covariances are not
statistically differentiable from zero (which could occur if LD and additive
genetic variance decay). However, the all other temporal covariances do not fit
the pattern we would expect under genome-wide fluctuating selection.

@@

  \caption{\textbf{A}: Scatterplot of the original unadjusted p-values from
    \textcite{Bergland2014-ij} and the p-values from our reanalysis of the same
    data using the same statistical methods; the minor discrepancy is likely
    due to software version differences. \textbf{B}: The histograms of the
    p-values of our reanalysis and the original \textcite{Bergland2014-ij}
    data; again the minor discrepancy is likely due to software differences.
    Overall, our implementation of Bergland et al.'s statistical methods
    produces results very close to the original analysis.}

@@

It is unclear how strong the fluctuations would have to be to generate a
genome-wide average signal of fluctuating selection from temporal covariances.
For example, many loci could still show a signal of fluctuating selection, but
the average signal could be overwhelmed by other signals of other selection. To
investigate whether there was an genome-wide excess of loci showing evidence of
fluctuating selection we reanalyzed the data of \parencite{Bergland2014-ij}
using the same seasonal fluctuating model as the original paper. This model is
a Binomial logit-linked GLM fit per-locus, where the Spring/Fall seasons are
encoded as a dummy variable, and are regressed on the frequency data. We use
the same binomial weighting procedure as \textcite{Bergland2014-ij}, where the
weights are determined by the effective number of chromosomes, $N_{eff} = (2
n_t d_t - 1) / (2 n_t + d_t)$ ($n_t$ and $d_t$ are the number of diploid
individuals and the read depth at timepoint $t$, respectively). We fit this
model on all loci marked as used in the VCF provided with the
\textcite{Bergland2014-ij} study (doi:10.5061/dryad.v883p). Overall, our
p-values for the Wald test for each locus closely match those of the original
paper (Pearson correlation coefficient 0.98, p-value < $2.2 \times 10^{-16}$;
see Supplementary Figure \ref{suppfig:bergland-pvalue-comparison} A), and the
histograms of the p-values are nearly identical (Supplementary Figure
\ref{suppfig:bergland-pvalue-comparison} B). \textcite{Bergland2014-ij} find
loci with a significant association with season after a Benjamini and Hochberg
FDR p-value adjustment \parencite{Benjamini1995-jy}, however, the null
hypothesis of the Wald test does not give us an idea of the expected number of
variants that may spuriously fit the pattern of seasonal fluctuating selection
as it does not account for genetic drift or other forms of hitchhiking.

@@

\begin{figure}[!ht]
  \centering
  \includegraphics[width=\textwidth]{figures/bergland-combined-hists.pdf}

  \caption{\textbf{A}: Histogram of original \textcite{Bergland2014-ij}
    seasonal p-values and p-values creating by randomly permuting the seasons
    at each locus. \textbf{B}: Histogram of original \textcite{Bergland2014-ij}
    p-values alongside all unique permutations (ignoring symmetries that lead
  to identical p-values).}

  \label{suppfig:bergland-pvalue-hist}
\end{figure}
